Enforced Invariant Confluence

### 2020-01-13 doc/the_vldb_journal_paper/sections/merge_reduction.tex
Added tech report sections of paper.

\begin{techreport}
  The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
  induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
  replace any subexpression that merges two merge-free subexpressions with an
  equivalent merge-free reachable subexpression (which we can do because $O$ is
  merge-reducible). We repeat this process until $e$ has been completely
  replaced with an equivalent merge-free reachable expression $e'$. Because
  $I(s_0)$ and because our system model only executes transactions that
  preserve the invariant, $e'$ (and hence $e$) is guaranteed to satisfy the
  invariant. Thus, all reachable states satisfy the invariant, so $O$ is
  \invariantconfluent{}.
\end{techreport}

@@

\begin{theorem}\thmlabel{LatticeProperty}
  Given an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$, if the following criteria are met,
  then $O$ is \sTIconfluent{}.
  \begin{enumerate}
    \item
      $O$ is a semilattice.

@@

\begin{techreport}
  The proof of \thmref{LatticeProperty} involves proving that if criteria (1)
  through (4) are met, then $O$ is \sTImergereducible{}. From this, it's
  immediate that $O$ is \sTIconfluent{} by
  \thmref{ReducibilityImpliesIconfluence}. To show that $O$ is
  \sTImergereducible, we consider two merge-free reachable expressions
  \[
    e_1 = t_1(\ldots(t_n(s_0))\ldots)
    \quad\text{and}\quad
    e_2 = u_1(\ldots(u_m(s_0))\ldots)
  \]
  Then, we inductively use criterion (3) with criteria (1) and (2) to show that
  \[
    e_3 = t_1(\ldots(
    t_n(u_1(\ldots(u_m(s_0))\ldots))
    )\ldots)
  \]
  is reachable and evaluates to $e_1 \join e_2$.
\end{techreport}

\thmref{IclosureImpliesIconfluence} states that \invariantclosure{} is a
sufficient condition for \invariantconfluence{}, and \thmref{LatticeProperty}
states that criteria (1) -- (4) are sufficient conditions for
\invariantconfluence{}. How do these sufficient conditions relate to one
another?  Clearly, not all \invariantclosed{} objects are semilattices, so
\invariantclosure{} does not imply criteria (1) -- (4). Conversely, there are
some objects which satisfy criteria (1) -- (4) which are not
\invariantclosed{}. Here's one example.

\begin{example}\examplelabel{TwoSets}
  Let $O = (\mathcal{P}(\nats), \cup)$ where $\mathcal{P}(\nats)$ is the power
  set of the natural numbers. Our start state $s_0 = \set{0}$ is the set of
  $0$. Let $t_Y(X) = X \cup Y$ be the transaction which unions $Y$ with its
  argument $X$. Our set $T = \setst{t_Y}{Y \subseteq \nats}$ of transactions
  consists of all possible $t_Y$. Our invariant $I$ consists of all non-empty
  sets $X$ that contain only even or only odd elements. That is, $I = \setst{X
  \subseteq 2\nats}{X \neq \emptyset} \cup \setst{X \subseteq 2\nats + 1}{X
  \neq \emptyset}$.

@@

  \begin{techreport}

    Here's why criterion (3) is satisfied. If $s$ is an arbitrary state that
    satisfies $I$, then it is non-empty and contains, without loss of generality,
    only even integers. If $t_1$ and $t_2$ are arbitrary transactions such that
    $I(t_1(s))$ and $I(t_2(s))$, then $t_1(s)$ and $t_1(s)$ are also non-empty
    and contain only even integers. Thus, $t_1(s) \cup t_2(s)$ is clearly
    non-empty and contains only even integers, so $I(t_1(s) \join t_2(s))$.
  \end{techreport}
\end{example}

@@

\Invariantclosure{} is not necessary for \invariantconfluence{} because it
fails to incorporate any notion of reachability. Criteria (1) -- (4) are also
unnecessary, but they can be used to prove that some non-\invariantclosed{}
objects are \invariantconfluent{} because the criteria \emph{do} incorporate
notions of reachability. In particular, criterion (3) is a slight variant of
\invariantclosure{}; it also states that invariant satisfying states should be
closed under merge. The fundamental difference is that criterion (3) restricts
its attention to the merge of two states which are \emph{reachable} from a
common ancestor state.

>>>>

The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
replace any subexpression that merges two merge-free subexpressions with an
equivalent merge-free reachable subexpression (which we can do because $O$ is
merge-reducible). We repeat this process until $e$ has been completely replaced
with an equivalent merge-free reachable expression $e'$. Because $I(s_0)$ and
because our system model only executes transactions that preserve the
invariant, $e'$ (and hence $e$) is guaranteed to satisfy the invariant. Thus,
all reachable states satisfy the invariant, so $O$ is \invariantconfluent{}.

@@

\begin{theorem}\thmlabel{LatticeProperty}
  Given an object $O = (S, \join)$, a start state $s_0 \in S$, a set of
  transactions $T$, and an invariant $I$, if the following criteria are met,
  then $O$ is \sTImergereducible{} (and therefore \sTIconfluent{}).
  \begin{enumerate}
    \item
      $O$ is a join-semilattice.

@@

To show that $O$ is \sTImergereducible, we consider two merge-free reachable
expressions
\[
  e_1 = t_1(\ldots(t_n(s_0))\ldots)
  \quad\text{and}\quad
  e_2 = u_1(\ldots(u_m(s_0))\ldots)
\]
Then, we inductively use criterion (3) with criteria (1) and (2) to show that
\[
  e_3 = t_1(\ldots(
  t_n(u_1(\ldots(u_m(s_0))\ldots))
  )\ldots)
\]
is reachable and evaluates to $e_1 \join e_2$.

\thmref{IclosureImpliesIconfluence} states that \invariantclosure{} is a
sufficient condition for \invariantconfluence{}, and \thmref{LatticeProperty}
states that criteria (1) -- (4) are sufficient conditions for
\invariantconfluence{}. How do these sufficient conditions relate to one
another?  Clearly, not all \invariantclosed{} objects are semilattices, so
\invariantclosure{} does not imply criteria (1) -- (4). Conversely, there are
some objects that satisfy criteria (1) -- (4) that are not
\invariantclosed{}. Here's one example.

\begin{example}\examplelabel{TwoSets}
  Let $O = (\mathcal{P}(\nats), \cup)$ where $\mathcal{P}(\nats)$ is the power
  set of the natural numbers. Our start state $s_0 = \set{0}$ is the set of
  $0$. Let $t_Y(X) = X \cup Y$ be the transaction that unions $Y$ with its
  argument $X$. Our set $T = \setst{t_Y}{Y \subseteq \nats}$ of transactions
  consists of all possible $t_Y$. Our invariant $I$ consists of all non-empty
  sets $X$ that contain only even or only odd elements. That is, $I = \setst{X
  \subseteq 2\nats}{X \neq \emptyset} \cup \setst{X \subseteq 2\nats + 1}{X
  \neq \emptyset}$.

@@

  Here's why criterion (3) is satisfied. If $s$ is an arbitrary state that
  satisfies $I$, then it is non-empty and contains, without loss of generality,
  only even integers. If $t_1$ and $t_2$ are arbitrary transactions such that
  $I(t_1(s))$ and $I(t_2(s))$, then $t_1(s)$ and $t_1(s)$ are also non-empty
  and contain only even integers. Thus, $t_1(s) \cup t_2(s)$ is clearly
  non-empty and contains only even integers, so $I(t_1(s) \join t_2(s))$.
\end{example}

@@

\Invariantclosure{} is not necessary for \invariantconfluence{} because it
fails to incorporate any notion of reachability. Criteria (1) -- (4) are also
unnecessary, but they can be used to prove that some non-\invariantclosed{}
objects are \invariantconfluent{} because the criteria \emph{do} incorporate
notions of reachability. In particular, criterion (3) is a slight variant of
\invariantclosure{}; it also states that invariant satisfying states should be
closed under merge. The fundamental difference is that criterion (3) restricts
its attention to the merge of two states that are \emph{reachable} from a
common ancestor state.
